A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-08 tex/codatadp.tex
tuned intro

@@

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have an important
role to play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
execution traces.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values of codatatypes
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all well-typed finite and infinite ground constructor
terms, and only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that faithfully
models the natural numbers extended with an
infinity $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
codatatype value.

Datatypes and codatatypes are an integral part of most popular proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, more and more
automatic theorem provers have been extended with dedicated reasoning facilities
for datatypes. The SMT-LIB~2 \cite{barrett-et-al-2010} syntax, implemented by
most SMT (satisfiability modulo theories) solvers, includes a theory of
datatypes.

In this \thewordpaper, we introduce a unified decision procedure for ground and
universal problems involving both datatypes and codatatypes. The procedure is
described abstractly as a calculus (Section~\ref{sec:the-calculus}) and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, ours is the first decision procedure for codatatypes.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. There are, however, at least three important
differences:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

@@

\cheat
\item
\emph{It must be possible to express cyclic values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic ($\omega$-regular)
values internally. For example,
the term $\const{SCons}(1,\, \mu s.\; \const{SCons}(0,\, \const{SCons}(9,\, s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\kern0.1em. %%% TYPESETTING: kern
\end{itemize}

@@

The cyclicity and bisimilarity rules are
needed to ensure completeness on ground problems and the absence of spurious
models in the context of finite model finding.
They cannot be replaced by finite
axiomatizations, so they naturally belong in a decision procedure. 
%    * in particular, acyclicity and bisimilarity are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
For the other (co)datatype properties---the injectivity, distinctness, and
exhaustiveness of constructors and the selector laws---it is
widely recognized that decision procedures are more efficient than
axiomatizations.
%axiom.

\begin{paper}
The proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least tolerated it, both of which we are thankful for.
%
Andreas Lochbihler and Peter Gammie shared their private codatatype-based
theories with us so we could include them in the benchmarks.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).

>>>>

\newcommand\afterLdots{\kern.1em} %% TYPESETTING

@@

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have an important
role to play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all depend on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground constructor
terms, and only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinity value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
codatatype value $\infty$.

Datatypes and codatatypes are an integral part of most popular proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, some
automatic theorem provers have been extended with dedicated reasoning facilities
for datatypes. The SMT-LIB~2 \cite{barrett-et-al-2010} syntax, implemented by
most SMT (satisfiability modulo theories) solvers, includes a theory of
datatypes.

In this \thewordpaper, we introduce a unified decision procedure for ground and
universal problems involving both datatypes and codatatypes. The procedure is
described abstractly as a calculus (Section~\ref{sec:the-calculus}) and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, our procedure is the first of its kind for the theory of
codatatypes.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. There are, however, at least three important
differences:
%
\begin{itemize}
\item
\emph{Codatatypes need not be well-founded.}
For example, the type
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
%
of infinite sequences, or streams, over an element type $\tau$ is allowed; the
corresponding datatype would be rejected as non-well-founded.

@@

\cheat
\item
\emph{It must be possible to express cyclic }(\emph{$\omega$-regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic
values internally. For example,
the term $\const{SCons}(1,\, \mu s.\; \const{SCons}(0,\, \const{SCons}(9,\, s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\afterLdots.
\end{itemize}

@@

The cyclicity and bisimilarity rules are
needed to ensure completeness on ground problems and the absence of spurious
models in the context of finite model finding.
They cannot be replaced by finite
axiomatizations, so they naturally belong in a decision procedure. 
%    * in particular, acyclicity and bisimilarity are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
For the other (co)datatype properties---the injectivity, distinctness, and
exhaustiveness of constructors and the selector laws---it is
widely recognized that decision procedures can be more efficient than
axiomatizations.
%axiom.

\begin{paper}
The proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least tolerated it, both of which we are thankful for.
%
Andrei Popescu helped clarify our thoughts about axiomatizing codatatypes.
Andreas Lochbihler and Peter Gammie shared their private codatatype-based
theories with us so we could include them in the benchmarks.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
